One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2021-09-03 paper.tex
edits up to 3.4

This section describes the statistical model for the spatial pattern of observed animal locations. We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda$, i.e.  the intensity at location $\bs$ is a random variable denoted $\lambda(\bs)$.  The log-Gaussian Cox process is a flexible model that can include spatial covariates to model the mean intensity as well a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.  To account for the imperfect detection of points we specify a thinning probability function $g(\bs) = \mathbb{P}(\text{a point at $\bs$ is detected } |\text{ a point is at $\bs$})$. A key property of the log-Gaussian Cox process is that the point pattern of \textit{detected} points itself follows a log-Gaussian Cox process with intensity process $\tl(\bs) := \lambda(\bs)g(\bs)$.  Using a log link this implies that $\log g(\bs)$ is included as an additive term to the model for $\log \lambda(\bs)$.

@@

A point transect distance sampling survey consists of a set of $K$ point transects.   We denote $k$th subset of space covered by the $k$th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.  For simplicity, we assume that all point transects are non-overlapping discs with equal radius $W$, as is the case in the HFBS survey.  It would be straightforward to relax this assumption.  The probability of observing a point at location $\bs \in \Omega_k$ given it is at $\bs \in \Omega_k$ we denote $g_k(\bs)$.  The probability of observing a point outside the surveyed region is zero.
Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single thinning probability function $g_k$.  For example, letting $\bs_k$ denote the centre of transect $\Omega_k$, the half-normal thinning probability function for points in $\Omega_k$ is $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.  The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation.  The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ for $\bs \in \Omega_k$.

@@

\sloppy Here we show how to approximate the thinned log Gaussian Cox process model with a modified Poisson likelihood.  The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.  Replacing the integral with a weighted sum allows the log-likelihood to be approximated as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.

To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.   In general, the thinning function depends only on $r$ and not on $\theta$.  We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$. The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.  This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
\begin{align*}
	\int_0^{2\pi} \lambda(s_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
	&= 2\pi \lambda(s_k)g_k(r)
\end{align*}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).  Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.   For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.  This gives
\begin{equation*}
	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
\end{equation*}
where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.

To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly, let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.  Then the approximate log-likelihood is
\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
\end{equation}
This approximation can be expressed as a modified Poisson likelihood.  To see this let $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood becomes
\begin{equation}
\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i)
\end{equation}
where $C$ is a constant.  This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.  This is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.

>>>>

This section describes the statistical model for the spatial pattern of observed animal locations. We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda$, i.e.  the intensity at location $\bs$ is a random variable denoted $\lambda(\bs)$.  The log-Gaussian Cox process is a flexible model that can include spatial covariates to model the mean intensity as well a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.  To account for the imperfect detection of points we specify a thinning probability function $g(\bs) = \mathbb{P}(\text{a point at $\bs$ is detected } |\text{ a point is at $\bs$})$. A key property of the log-Gaussian Cox process is that the point pattern of \textit{detected} points itself follows a log-Gaussian Cox process with intensity process $\tl(\bs) := \lambda(\bs)g(\bs)$.  Using a log link $\log g(\bs)$ is therefore included in the model as an additive component to the model for $\log \lambda(\bs)$.

@@

A point transect distance sampling survey consists of a set of $K$ point transects.   We denote $k$th subset of space covered by the $k$th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.  For simplicity, we assume that all point transects are non-overlapping disks with equal radius $W$, as is the case in the HFBS survey.  It would be straightforward to relax this assumption.  The probability of observing a point at location $\bs \in \Omega_k$ given it is at $\bs \in \Omega_k$ we denote $g_k(\bs)$.  The probability of observing a point outside the surveyed region is zero.
Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single thinning probability function $g_k$.  For example, letting $\bs_k$ denote the centre of transect $\Omega_k$, the half-normal thinning probability function for points in $\Omega_k$ is $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.  The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation.  The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ for $\bs \in \Omega_k$.

@@

\sloppy The likelihood \eqref{lgcp-likelihood} contains an integral that cannot be solved analytically.  A common approach for fitting log-Gaussian Cox processes in practice is to approximate the likelihood by replacing the integral with a weighted sum.  In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.  We follow closely with the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process but adapted for the point transect distance sampling case.

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the transect.  It should be clear from context which is intended by $\bs_k$.   The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.

The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.  This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.  Therefore
\begin{align}
\label{eq-linear-intensity}
	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
	&= 2\pi \lambda(\bs_k)g_k(r)
\end{align}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the GMRF spatial effect (Section \ref{sec-gmrf}).

It follows that $\int \lambda(\bs)g(\bs) \mathrm{d}\bs = \sum_{k=1}^K 2\pi \lambda(\bs_k) \int_0^W r g_k(r)\mathrm{d}r$ by applying the change of variables formula for integration and substituting in \eqref{eq-linear-intensity}.  Note that this integral now only involves the one dimensional integral with respect to $r$.  For each sampling unit we approximate this integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.  This gives
\begin{equation*}
	\int\displaylimits_{\bs \in \Omega} \lambda(\bs)g(\bs)\mathrm{d}\bs \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(\bs_{kj}) \;\;\; ,
\end{equation*}
where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(\bs_{kj}) = \lambda(\bs_k) g_k(r_{kj})$.

To simplify notation below we gather all integration weights into the vectors $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.  Similarly, let $\tl_k = (\tl(\bs_{k1}), \ldots, \tl(\bs_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$, the thinned intensity evaluated at each integration location, and $\tl_{obs} = (\tl(\bs_1), \ldots, \tl(\bs_n))^\intercal$, the thinned intensity evaluated at each observation location.  Then the approximate log-likelihood can be written as
\begin{equation}
\label{approx-log-likelihood}
	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
\end{equation}
This approximate likelihood can be expressed as a modified Poisson likelihood by letting $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and constructing a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.  Then the approximate likelihood is
\begin{equation}
\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i)
\end{equation}
where $C$ is a constant.  This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.

The approach taken here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
